---
title: R/lmmlite user guide
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R/lmmlite user guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

[R/lmmlite](http://kbroman.org/lmmlite) is an R package for fitting linear
mixed models in the context of genome-wide association studies (GWAS)
and quantitative trait locus (QTL) mapping. At present, **it is not intended for
"production" use.** I wrote this to try to get a better
understanding of how to fit these models, and I've not yet put any
effort into checking arguments and omitting cases with missing values.

The code closely follows [Nick Furlotte](http://whatmind.com)'s
[pylmm](https://github.com/nickFurlotte/pylmm).

The sort of model I want to fit is
$y = X\beta + \epsilon$ where $\epsilon$ is multivariate normal with
mean 0 and variance $\sigma^2_g K + \sigma^2_e I$, where $K$ is a
known kinship matrix and $I$ is the identity. In pylmm, the focus is
on $\sigma^2 = \sigma^2_g + \sigma^2_e$ and the heritability
$h^2 = \sigma^2_g / \sigma^2$.

```{r load_data, echo=FALSE}
library(lmmlite)
data(recla)
```

We'll focus here on the `recla` dataset that I include with the
package. This is data on a set of diversity outcross mice, from
[Recla et al. (2014)](http://www.ncbi.nlm.nih.gov/pubmed/24700285) and
[Logan et al. (2013)](http://www.ncbi.nlm.nih.gov/pubmed/23433259). The data are
originally from the
[QTL Archive/Mouse Phenotype Database](http://phenome.jax.org/db/q?rtn=projects/projdet&reqprojid=285)
and I've also
[placed them online](https://github.com/kbroman/qtl2data/tree/master/DO_Recla)
in the format used by [R/qtl2](http://kbroman.org/qtl2). I used
[qtl2geno](https://github.com/rqtl/qtl2geno) to calculate genotype
probabilities from the MUGA SNP array data. The data include
`r nrow(recla$kinship)` mice, and form a list with three components:
the estimated kinship matrix, a matrix with `r ncol(recla$pheno)`
phenotypes, and a covariate matrix that contains an intercept (all
1's) and sex (0 for female and 1 for male).

```{r show_load_data, eval=FALSE}
library(lmmlite)
data(recla)
```

The package includes two implementations of the fit of linear mixed models
(LMMs): plain R, and C++ (using the
[Rcpp](https://github.com/RcppCore/Rcpp) and
[RcppEigen](https://github.com/RcppCore/RcppEigen) packages, the
latter being an interface to the [Eigen](http://eigen.tuxfamily.org/)
C++ linear algebra library).

R/lmmlite includes just four functions: `eigen_rotation()`,
`getMLsoln()`, `calcLL()`, and `fitLMM()`. In each case, use
`use_cpp=FALSE` to use the plain R implementation and `use_cpp=TRUE`
(the default) to use the C++ implementation.

The first function, `eigen_rotation()`, calculates the eigen
decomposition of the kinship matrix and "rotates" the phenotypes and
covariates by pre-multiplying them with the tranpose of the matrix of
eigenvectors. This rotation greatly simplifies later calculations by
turning a general least squares problem into a weighted least squares
problem.

Care must be taken with missing values in the phenotypes and
covariates. In practice, one must work with batches of phenotypes that
have a common missing data pattern, and first remove individuals with
missing values. Here, we'll focus on the first phenotype.

```{r eigen_rotation}
e <- eigen_rotation(recla$kinship, recla$pheno[,1], recla$covar)
```

The output is a list containing four components: `Kva` is the vector
of eigenvalues of the kinship matrix, `Kva_t` is the transposed matrix
of eigenvectors, `y` is the rotated phenotype matrix, and `X` is the rotated
covariate matrix.

The next function `getMLsoln()` will estimates the coefficients
($\beta$) and residual variance ($\sigma^2 = \sigma^2_g + \sigma^2_e$)
for a given value of the heritability ($h^2 = \sigma^2_g / \sigma^2$).
It takes, as input, a single value for the heritability, and then
`Kva`, `y`, and `X`, as returned from `eigen_rotation()`.  A further
argument `reml`, indicates whether REML will later be used, in which
case the log determinant of the matrix $X' W X$, where $W$ is a
diagonal matrix of weights, is calculated.

```{r getMLsoln}
ml_soln <- getMLsoln(0.5, e$Kva, e$y, e$X)
```

The result is a list containing `beta` and `sigmasq`, with the
residual sum of squares (RSS) and log det($X' W X$) included as
attributes.

The function `calcLL()` calculates the log likelihood for a given
value of the heritability. (It can also take a vector of
heritabilities.) It takes the same arguments as `getMLsoln()`, and in
practice one would probably never call `getMLsoln()` directly.

```{r calcLL}
hsq <- seq(0, 1, by=0.01)
ll <- calcLL(hsq, e$Kva, e$y, e$X)
```

We can plot the results as follows.

```{r plot_ll, fig.width=7, fig.height=4}
plot(hsq, ll, type="l", lwd=2, las=1,
     xlab="heritability", ylab="log likelihood")
```

If you call `calcLL()` with a single heritability value, the results
include the estimates of `beta` and `sigmasq` as attributes.

```{r calcLL_one}
one_ll <- calcLL(0.5, e$Kva, e$y, e$X)
attr(one_ll, "beta")
attr(one_ll, "sigmasq")
```

The final function, `fitLMM()`, optimizes the log likelihood to
estimate the heritability. The result is a list containing `beta`,
`sigmasq`, `hsq`, `sigmasq_g`, `sigmasq_e`, and `loglik`.

```{r fitLMM}
(out <- fitLMM(e$Kva, e$y, e$X))
```

The `fitLMM()` function includes an argument `check_boundary`; if
`TRUE` (the default), the 0 and 1 boundaries are checked explicitly, in seeking to
estimate the heritability.
